---
title: "CAP5771 Final Project"
author: "Quentin Goss `quentingoss0323@floridapoly.edu`"
subtitle: "Exploration of Street Network Node and Edge Features."
bibliography: "CAP5771-final.bib"
cls: "IEEEtran.cls"
output:
    html_document:
        df_print: paged
---
<hr>
## *Part 1 - Introduction*

**<u>Research Question</u>:** What do the features of street network nodes and vertices tell us about a street network, which features are dominant, and how can these features be used to make a general classification of a street network?

***Abstract*--One of the many steps in modeling transportation systems, is understanding the street network and the characteristics of the roads and intersections thereof. In this paper, we utilize R to perform principle component analysis and association rule learning to explore and classify the characteristics of a street network spatially.**

***Index Terms*--transportation systems, street networks, association rule learning, principle component analysis**

The transportation system (TS) is a complex arrangement of roads, highways, bridges, intersections, and infrastructure which traditionally provides the means of moving people and cargo from one place in the system to another in the least amount of time. Every TS is unique, and also imperfect since the layout of a TS is gradually implemented over ten or hundreds of years. During the expansion of a TS, it is unreasonable or impossible to predict the perfect expansion layout for the current moment as well for all potential futures. Often times, a government committee creates TS models to plan for TS expansion in the near future such as the 5 Year Consolidated Plan created by the Polk County Board of County Commissioners each year @PCBOCC2016.

The street network (SN) is the skeleton of a TS, as it consists of the paths that vehicles in a TS travel along to traverse the TS. These SNs may be represented as graphs $G = \{E,V\}$ which are a combination of intersections $V$ and road segments between intersections $E$. Realistic SNs, such those obtained from Open Street Maps (OSM) @OpenStreetMap2019 are very detailed, including data about speed, highway type, merging rules at an intersection,priorities for edges to break same-arrival ties at intersections, etc.

Many TS modeling tools utilize SNs and the OSM SNs such as OSMnx for analysis and model construction @BOEING2017, CrowdSenSim for building crowdsensing models @FIANDRINO2017, and by @VITELLO2018 in their tool for pedestrian mobility modeling.

We will be using a SN retrieved from OSM, of which the drivable roads are filtered and that is converted into SUMO SN format, which is easier to work with.

Edges may contain data such as speed, type, priority, and name, along with graph information such as the vertex id that it begins and ends at. The nodes (vertices) are simpler, they contain spatial information, and a type, which are tie-breaking rules when two vehicles enter the vertex, which decides a vehicle to move first when they arrive at an intersection at the same time. For our proposed approach, we will first pre-process the SN data using Python to format it into CSV (comma separated value) format. Then, we will utilize R to explore the SN data, performing principle component analysis (PCA) to determine which edge properties are most important to defining the edges of a SN, and also association rule learning using the Apriori algorithm to create association rules that can be used to model a similar SN, or to be used to enhance the quality of a dimension reduced SN.

We will use the term **"nodes"** in place of **"vertices"** for the remainder of this report as it is convention when working with OSM and SUMO SN data.

<hr>

## *Part 2 - Exploratory Data Analysis (EDA)*

We have selected a $5000m^2$ SN of Orlando to explore. We will perform EDA on the **nodes** first, then on the **edges**.

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(dplyr)
library(cowplot)
```



#### Nodes

```{r}
node.csv <- "https://raw.githubusercontent.com/QuentinGoss/CAP5771-Final-Project/master/data/node.csv"
nodes.raw <- read.csv(node.csv)
```

The **nodes** are basic. In most cases with SUMO SNs, only four of the fields are utilizied. For brevity those four are as follows:

Feature | Description
------- | -----------
id      | Unique Identifier
x       | x coordinate
y       | y coordinate
type    | Rules for resolving same-time arrival conflicts

The **type** options are self exclamatory -- they may be any of the following: priority, traffic_light, right_before_left, unregulated, allway_stop, rail_signal, zipper, traffic_light_right_on_red, and rail_crossing.

```{r}
paste0('Rows: ',nrow(nodes.raw))
summary(nodes.raw)
```

<!-- We will remove the columns that have no data and keep **id**, **x**, **y**, and **type**. -->

<!-- ```{r} -->
<!-- nodes <- nodes.raw[, c('id','x','y','type')] -->
<!-- head(nodes) -->
<!-- ``` -->

<!-- Let's plot the intersections and color them by type. We exclude nodes of **type** *priority* since we anticipate that most of the nodes are going to be *priority*, which is hard to visualize. -->

<!-- ```{r} -->
<!-- nodes.nopriority <- nodes[ which(nodes$type != 'priority'),] -->
<!-- ggplot(nodes.nopriority, aes(x=x,y=y,color=type)) + -->
<!--     geom_point() -->
<!-- ``` -->

We can't do anthing with the node data related to our research question because:
1. The node type is not numerical
2. The node type is the only feature aside from **x**,**y** and **id**, which are not relevant to the research question.
3. This SN doesn't have any other node features of importance.

Therefore, after EDA of the node data, we reach conclusion 1.

**<u>Conclusion 1</u>:** We may not apply principle component analysis to the node data, nor may we apply association rules learning to the node data because there is only one relavant categorical feature. Therefore, we do not consider node data to be considered in the classfication of the SN.

#### Edges

```{r}
edge.csv <- "https://raw.githubusercontent.com/QuentinGoss/CAP5771-Final-Project/master/data/edge.csv"
edges.raw <- read.csv(edge.csv)
```

The edges of the SN seem to be much more promising to describe the SN than the nodes with 24 features.

Feature | Description
------- | -----------
id | Edge ID
from | Starting node ID.
to | Ending node ID
type | Classification of road (i.e Interstates and County Roads)
numLanes | The amount of lanes.
speed | Speed limit (meters/second)
spreadType | Which lane vehicles default to when entering the edge \{"right","center"\}
width | Road width (meters)
name | Name of the real-life road that the edge corresponds with. A blank name means offramp or trivial connector road.
endOffset | The distance the edge if differnt than the (x,y) of **to**.
sidewalkWidth | Width of the sidewalk
length | Length of the edge (meters)

The remaining columns are the vehicle types that are allowed to travel on the edge.

<!-- ```{r} -->
<!-- head(edges.raw) -->
<!-- ``` -->

Now we take a subset of the data, and clean it up. As stated in the research question, we will be performing Principle Component Analysis (PCA) and Association Rules Learning (ARL), therefore we should select quantitative features for PCA and categorical features for ARL.

***Columns that will be pruned:*** <br>
**id**, **from**, and **to** consist of features that are unique to this SN and no other. These are defined by OSM, and are unique to itself from the set of every edge in the world. **width**, **endOffset** and **sidewalkWidth** are full of ''. Finally we prune the allowed vehicle type columns **emergency** $\cdots$ **moped** because the values are uniform across the entire dataset.

***Rows that will be pruned:*** <br>
The 269 rows that have '' as the **name** will be trimmed. This will leave 7074 edges, a significant amount.

***Values that will be modified:***<br>
`NA` values of **spreadType** will be changed to "right", the default for SUMO edges.

***Note:***
We have opted to keep the **name** feature, since road names are often commonplace, as is known from the ***Census and You***, a document by the @common-road-names.

<!-- ```{r} -->
<!-- paste0('rows: ',nrow(edges.raw)) -->
<!-- summary(edges.raw) -->
<!-- ``` -->

The code for cleaning the data.

```{r}
# Trim the decided columns
edges <- edges.raw[,c('type','numLanes','speed','priority','spreadType','name','length')]

# Trim rows where name = NA
edges <- edges[which(edges$name != ''),]

# Change spreadType '' to 'right'
levels(edges$spreadType)[levels(edges$spreadType) == ''] <- 'right'

summary(edges)
```

The **speed**, **priority**, and **numLanes** features have a significant quality that makes them both categorical and numerical. That is, that all of 7079 samples fit into a small number of buckets.

```{r}
type.vs.speed <- ggplot(edges, aes(x=speed, y=type)) + geom_point()
priority.vs.numlanes <- ggplot(edges, aes(x=priority, y=numLanes)) + geom_point()
plot_grid(type.vs.speed, priority.vs.numlanes, labels= "AUTO")
```


#### *EDA Conclusions*

During EDA we have realized that the node data does not have enough features to perform PCA or ARL to reach a generalization or classification of the SN after the **id**, **x**, **y**, and empty columns are pruned, leaving one feature. 

However, from the 24 features of the raw edge data, we have identified 7079 rows and 7 features. 4/7 of the features; **numLanes**, **speed**, **priority**, and **length** are numeric, thus applicable to perform PCA on. 6/7 of the features; **type**, **spreadType**, **name**, **speed**, **numLanes**, and **priority** are categorical, thus applicable to perform ARL on.

Even without the node data, the edge data satisfies the PCA and ARL applicability conditions of the research question, therefore we may proceed to the methodology portion.

<hr>

## *Part 3 - Methods*

#### *Principle Component Analysis*

```{r message=FALSE, warning=FALSE}
library(caret)
library(ggfortify)
```


```{r}
edges.numeric <- edges[,c('numLanes','speed','priority','length')] 
pca.edges <- prcomp(edges.numeric, scale = TRUE)
summary(pca.edges)
```

```{r}
autoplot(pca.edges, loadings.colour = 'blue', loadings = TRUE, loadings.label = TRUE, loadings.label.size = 5)
```

```{r message=FALSE}
edges.numeric.std <- edges.numeric %>% 
    preProcess(method = c('center', 'scale')) %>% 
    predict(edges.numeric)
autoplot(cor(edges.numeric.std)) +
    labs(x = '', y = '', title = 'Correlation Plot')
```


#### *Association Rules Learning*

```{r}
edges.categorical <- edges[,c('type','spreadType','name','speed','numLanes','priority')]
```


<hr>

## *Part 4 - Conclusions.*

**<u>Conclusion 1</u>:** We may not apply principle component analysis to the node data, nor may we apply association rules learning to the node data because there is only one relavant categorical feature. Therefore, we do not consider node data to be considered in the classfication of the SN.

<hr>

## *Aknowledgements*
The street network data is partially derived from OSM data (c) OSM contributors and available at [https://www.openstreetmap.org/](https://www.openstreetmap.org/)

<hr>

## *References*